A conditional permutation-based approach to test confounder effect and center-bias in machine learning models

### 2021-08-11 manuscript.tex
polished

Obviously, a strong association between $\yhat$ and $\c$ might indicate that the model is biased; it's predictions are driven by the confounder rather than information about the target variable.
Testing unconditional independence between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confounder-bias in predictive modelling.
For instance, even if the null hypothesis $H0 : \yhat \independent \c$ is false, $\yhat$ might be only marginally dependent on $\c$, due to their dependence on $\y$. In other words, if the target variable $\y$ is associated to the confounder variable $\c$, a model that is completely blind to $\c$ (and, therefore, unbiased) might still provide outputs $\yhat$ that are significantly associated with $\c$.

 Below, I summarize how testing conditional independence \citep{dawid1979conditional} with conditional permutation test \citep{berrett2020conditional} allows evaluating confounder-bias in predictive modelling, without placing assumptions on the distribution of the model output, where normality is often violated \citep{garcia2009study, kristensen2017whole}.

@@

Depending on the research question, the influence of $\c$ on the model predictions $\yhat$ is subject to be kept at a negligible level or it is to be proven that, at least, the model is not completely driven by $\c$.

Our primary interest is, therefore, the \emph{conditionally independence} between $\yhat$ and $\c$ given $\y$ (written as $\yhat \independent \c | \y$) which, according to \cite{dawid1979conditional}, implies that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(\yhat|\y)\mathbb{P}(\c|\y)$. The test with this null hypothesis will be referred to as the \emph{partial confounder test}.

Alternatively, one might be interested in testing $\yhat \independent \c | \y$. We will refer to the corresponding test as the \emph{full confounder test}.

 Conditional independence - in its general form - lies at the heart of several fundamental concepts in statistics and  plays an increasingly important role in various fields of applied statistics, particularly in biomedical applications \citep{spirtes2000causation, peters2016causal, fiedler2011mediation, candes2016panning}. Recently, \cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
 Their "no free lunch" theorem implies that, without placing some assumptions on the joint distribution of $(\y, \yhat, \c)$, conditional Independence testing is effectively impossible. In other words, neither the full nor the partial confounder tests can be constructed so that - for all distributions - they provides a valid type I error control and, at the same time, a non-trivial statistical power.

@@

Nevertheless, as shown by \cite{berrett2020conditional}, conditional independence testing is possible with information only about the conditional distribution of two out of the three variables.
The conditional permutation test (CPT) of Berret and colleagues builds on the conditional randomization test of \cite{candes2016panning} and samples from a non-uniform distribution over the set of possible permutations $\pi$ of one of the variables, so that it incorporates available distributional information into permutation-based inference.

@@

As many related papers, the work of \cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is the predictor(s) (features), $Y$ is the target variable and $Z$ is a potential confounder. In this setting, testing the null hypothesis $X \independent Y | Z$ aims to determine, whether $X$ affects $Y$, when controlling for $Z$.
For instance, in genome-wide association studies, CPT can be used to determine whether a particular genetic variant $X$ affects a response $Y$ such as disease status or some other phenotype, even after controlling for the rest of the genome, encoded in $Z$.

@@

Within this setting, conditional independence testing and, specifically, the framework of conditional permutation testing allows investigating three different null hypotheses corresponding to the $(\y, \yhat, \c)$ triplet, as listed in Table \ref{tab:conditional-independence-cases}. Testing the null hypothesis $\y \independent \yhat | \c$ (option 1) investigates whether the model predictions are likely explainable solely with the confounder, i.e. whether the model is exclusively confounder-driven. Testing $\y \independent \c | \yhat$ (option 2) addresses the question whether the model captures all the variance in $c$ when predicting $y$. Testing the null hypothesis $\yhat \independent \c | \y$ (option 3) examines, whether the dependence of the model output on the confounder can likely be explained by the confounder's dependence on the target variable, i.e. whether there is any confounder bias in the model.

@@

Option 3, i.e. partial confounder testing is typically of interest when testing confounder bias of predictive models. Option 1, i.e. full confounder testing, might be also useful in diagnostics of predictive models, especially in the exploratory phase of model construction. Option 2 seems less appealing for model diagnostics and importantly, the CPT framework does not allow constructing a test which is non-parametric on $\yhat$.

Below, CPT is adapted for partial confounder testing (option 3). The formulation of the full confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.

\subsubsection*{The partial confounder test}

@@

$$\c_i^{(j)} \sim Q(\cdot|y_i) \ $$

where, $Q(.|y)$ denotes the conditional distribution of $\c$ given $\y=y$ and $j=1,\dots, m$ indexes the "copy" of $\c$ so that $\c^{(j)} = (c_1^{(j)}, \dots, c_n^{(j)})$ is a permutation of the original vector $\c = (c_1, \dots, c_n)$.

This mechanism creates copies $\c^{(1)}, \dots ,\c^{(m)}$ so that under the null hypothesis ($\yhat \independent \c | \y$), the triples $$(\y,\yhat,\c), (\y, \yhat, \c^{(1)}),\dots, (\y, \yhat, \c^{(m)})$$ are all identically distributed and exchangeable, and so are the 
$$T(\y,\yhat,\c), T(\y, \yhat, \c^{(1)}),\dots,T(\y, \yhat, \c^{(m)})$$
test statistics, as well.

Let us denote the permutation-based copies of

$$\c^{(j)} = \c_{\pi^{(j)}}, \quad \pi \in S_n$$

where $S_n$ denote the set of all permutations on the indices $\{1,\dots,n\}$ and $\c_\pi = (c_{\pi_1}, \dots, c_{\pi_n})$ the vector $\c$ with its elements reordered according to the permutation .

In the partial confoudner test, the $\c_{\pi^{(j)}}$ copies are drawn so that:

\begin{equation}
    \mathbb{P}(\pi^{(j)} = \pi | \y,\yhat,\c) = \frac{q^n(\c_\pi | \y)}{\sum_{\pi' \in S_n} q^n(\c_{\pi'} | \y)}
    \label{eq-pperm}
\end{equation}

that is, according to the $q^n(\cdot|\y) := q(\cdot | y_1) \dots q(\cdot|y_n)$ product density corresponding to the conditional distribution $Q(\cdot|\y)$. Note that Eq. \ref{eq-pperm} does not necessarily assume a continuous distribution.
For a verification of the valid type I error control of this approach refer to Theorem 1 in \citep{berrett2020conditional}.

@@

While the required permutations could be theoretically sampled with a simple Metropolis-Hastings algorithm that draws uniformly from $S_n$ at random, this way the acceptance ratio would be extremely low, even for moderate $n$ (except there is very low dependence of $\c$ on $\y$), resulting in slow mixing times. The partial confounder test can be, however, efficiently implemented with the parallelized pairwise Markov-Chain Monte Carlo sampler of \cite{berrett2020conditional} (Algorithm 1), that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data. In their Theorem 2, \cite{berrett2020conditional} verify that the resulting Markov Chain yields the desired stationary distribution, even if the number of steps is small.

\subsubsection*{Conditional log-likelihood matrix}

@@

with the assumption that the residual error follows a Gaussian distribution. If we write $\boldsymbol{\mu} = \alpha + \beta \y$ and $\sigma^2 = E(\boldsymbol{e}^2)$ then, for the conditional Normal distribution $ (\c|\y=y_i) \sim \mathcal{N}\{\mu_i, \sigma^2\}$ and the log-likelihood, that is to be used in Eq. \ref{eq-pperm}, can be computed as follows:

$$ log\mathcal{L}(\c|\y) = - \frac{n}{2} ln(2 \pi) -  \frac{n}{2} ln( \sigma^2) - \frac{1}{2 \sigma^2} \sum_{i=1}^n (c_i - \mu_i)^2 $$

@@

A straightforward choice of test statistic with the proposed Gaussian likelihood estimation is the coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification):

@@

for any $\pi \in S_n$

Notably, from the three options for conditional independence-based null hypotheses, enumerated in Table \ref{tab:conditional-independence-cases}, the proposed Gaussian likelihood estimation is possible for options 1 and 3, but not option 2, as the variable on which the independence is conditional must be always the predictor variable in Eq. \ref{eq:linreg}.

For options 1 and 3, the proposed simple Gaussian conditional likelihood estimation allows for deploying CPT without a large, independent sample to estimate the conditional distribution.

@@

'Partial' and 'full' confound testing was performed with default parameters (1000 permutations, 50 Markov-chain Monte Carlo steps) as implemented in version 0.9.0 of the package \emph{'mlconfound'}.

@@

Simulations based on normally distributed data justified that the proposed confounder tests - in line with theory - provide a valid type I error control (For detailed results, see first row and first column on Figure \ref{fig:sim-normal} for the 'partial' test on numerical data and Supplementary Figures \ref{fig:sim-ccc-full}, \ref{fig:sim-ccb-partial}, \ref{fig:sim-ccb-full}, \ref{fig:sim-bbc-partial}, \ref{fig:sim-bbc-full}, \ref{fig:sim-bbb-partial} and \ref{fig:sim-bbc-full} for the partial test categorical data and the full tests on both numeric and categorical data). The higher 95\% confidence limit for the type I error rates for one simulation case (with 100 repetitions) is 0.11. If the test is valid, five simulation instances are expected to pass this limit across 200 H0 simulations (all null hypotheses for all simulations depicted on the same figure). The number of simulation cases with type I error rate greater than 0.11 was always smaller then this expected value.
The results suggest, moreover, that the proposed tests are somewhat overly conservative if the model has a high predictive performance (type I error rates closer to 0 than expected).

Simulations with non-normal data showed that the proposed methods are remarkably robust to the violation of normality assumptions for the target variable $\y$ and the confounder $\c$ and, as expected - completely non-parametric regarding the distribution of the predictions $\yhat$). The six simulation cases of normality violation, shown on Figure \ref{fig:sim-non-normal} involved various \emph{sinh-arcsinh} distributions, form totally normal (Fisher's kurtosis: 0, skewness) to severely skewed and heavy tailed (kurtosis: 42, skewness: 6).
The results show that the violation of normality - as expected - increases the type I error rate, but the increase - depending on the application - might not be severely problematic up to a kurtosis and skewness values as high as 5 and 2 respectively. Furthermore, normality violation becomes less problematic in case of high predictive performance (high association between the predicted and observed values).

@@

When normality assumptions hold, the test generally possess a statistical power of high practical relevance. Power estimates for the 'partial' test (with normality) are shown on Figure \ref{fig:sim-normal}. Notably, with sample sizes as large as 1000, a confounder contributing to only 0.4-0.6\% to the variance of the predictions can already be successfully detected ($w_c = 0.05 - 0.07$) with a power of 60-90\%. With a sample size of 500, a confounder contributing to only 1.4-3\% to the variance of the predictions can be detected ($w_c = 0.1 - 0.14$) with a power greater than 80\% in most of the simulation cases. A sample size of 100 yields an approximate power of 60-100\% for a confounder-variance of 6-7\% ($w_c = 0.2 - 0.21$) in most of the cases. Finally, even with a low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%. If the confounder explains more than 15\% of variance, it is almost certainly detected with $n \geq 50$.

@@

Acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p=0.0014$ and $0.001$, respectively). The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
However, as revealed by the partial confounder test, the model was significantly biased both by age group and acquisition batch (both $p<0.0001$) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to larger predicted values.

After applying the confound mitigation approaches (feature regression or COMBAT) the partial confound test did not provide evidence of confounder bias anymore ($p < 0.05$ for all). Both feature regression and COMBAT slightly improved the predictive performance in the case of both confounders, with COMBAT providing better performances ($R^2=122$ and $121$ when applied to remove the effect of acquisition and age, respectively).

@@

Imaging center and in-scanner motion (normalized mean framewise displacement) were statistically significantly associated with ASD diagnosis ($R^2=0.027$ and $0.028$, respectively, $p<0.001$ for both). The model trained on the raw (unadjusted) connectivity features predicted diagnosis with a medium effect size ($R^2=0.095$, $ROC AUC = 0.71$, $p<0.001$).
However, as revealed by the partial confounder test the raw model was significantly biased both by age group and acquisition batch (both $p<0.0001$) with several sites (e.g. Carnegine Mellon University or Caltech) and higher motion being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to a higher probability of the model to assign the participant to the ASD group.

After applying the confound mitigation approaches to remove the effect of center (feature regression or COMBAT) the partial confound test did not provide evidence of center-bias anymore ($p < 0.05$ for both techniques). When trying to mitigate the effect of in-scanner motion, feature regression failed to remove the motion-bias from the model and, in fact, it introduced a paradoxical inverse dependence of the predictions on motion estimates. The partial confounder test successfully detected this bias ($p<0.0001$). Applying COMBAT to remove the effect of motion (by using binned motion estimates), in turn, effectively mitigated motion bias, as suggested by the partial confound test ($p=0.54$).

Both feature regression and COMBAT slightly improved the predictive performance when mitigating center-effects ($R^2=0.111$ and $0.132$ and $AUC=0.76$ and $0.75$, respectively). With both feature regression and COMBAT, the effort of mitigating motion effects happened at the cost of predictive performance ($R^2=0.009$ and $0.087$ and $AUC=0.59$ and $0.70$, respectively)

% tab:unconditional-pvals
\renewcommand{\arraystretch}{1.2}
\begin{table}
\centering
\begin{tabular}{lll|ll|ll|ll|ll} 
dataset & conf. & method & $R^2_{y, c}$ & $p_{y, c}$ & $R^2_{\hat{y}, c}$ & $p_{\hat{y}, c}$ & $R^2_{\hat{y}, y}$ & $p_{\hat{y}, y}$ & partial test & full test  \\
\hline
HCP & acq.  & raw      & 0.032 & <0.001 & 0.084 & <0.001  & 0.082 & <0.001 & <0.001 & <0.001 \\
    &              & f.reg.    & & &0.003 & 0.99 & 0.066  & <0.001  & 0.99 & <0.001 \\
    &              & COMBAT    & & &0.008 & 0.75 & 0.075  & <0.001  & 0.88 & <0.001\\
\hline
    & age   & raw       & 0.011 & 0.001  & 0.041 & <0.001  & 0.082  & <0.001 & <0.001 & <0.001 \\
    &       & f.reg.    & & &0.001 & 0.3 & 0.07  & <0.001  & 0.5 & <    0.001 \\
    &       & COMBAT    & & &0.004 & 0.055 & 0.076 & <0.001 & 0.12 & <0.001 \\
\hline
ABIDE   & center   & raw       & 0.005  & 0.203 &  0.169 & <0.001& 0.126     & <0.001 & <0.001 & <0.001 \\
        &          & f.reg.    &  & &  0.042 & 0.007 & 0.173     & <0.001 & 0.04 & <0.001 \\
        &          & COMBAT    &  & &  0.05 & 0.001 & 0.17     & <0.001 & 0.01 & <0.001 \\
\hline
        & motion   & raw       & 0.028 & <0.001 & 0.111    &  <0.001 & 0.126    & <0.001 & <0.001& <0.001 \\
        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & <0.001 & 0.09 \\
        &          & COMBAT    & & &  0.002 & 0.202 & 0.111     & <0.001 & 0.64 & <0.001 \\

@@

Confounder-bias can significantly harm the generalizability and biomedical validity of predictive models. Testing confounder-bias of predictive models can be achieved by considering the "confounder-problem" as assessing \emph{conditional independence} between the predicted values and either the target or the confounder variable. These two scenarios result in two different test, testing whether the model is partially or fully driven by the given confounder.
Importantly, in predictive modelling, the model predictions are often follow a non-normal distribution, thus the any test for confounder bias must be, at least partly, non-parametric.
However, as shown by in \citet{shah2020hardness} their the 'no free lunch' theorem, it is effectively impossible to establishing a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
Here I propose the use of the Markov-chain Monte-Carlo based conditional permutation test (CPT) of \cite{berrett2020conditional} to test confounder bias in predictive models. Adapting this approach for the confounder testing problem makes it possible to place assumptions about the joint distribution of only two out of the three variables involved; that is, it allows constructing both the partial and full confounder tests so that no assumption is made about the distribution of the model output. 
Here I propose a computationally effective parallel implementation of CPT that utilizes the parallel pairwise sampler of \cite{berrett2020conditional}, coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confounder is normal. The implementation provides a test that can be uniformly applied for any predictive model, given that realistic (e.g. cross-validated) model predictions can be obtained.

@@

Given the power and the robustness of the proposed tests, they might become a useful tool in diagnosing predictive models, not exclusively in the biomedical filed.

>>>>

Depending on the research question, the direct influence of $\c$ on the model predictions $\yhat$ is subject to be kept at a negligible level or it is to be proven that, at least, the model is not completely driven by $\c$.

Obviously, a strong association between $\yhat$ and $\c$ might indicate that the model is biased; it's predictions are driven by the confounder rather than information about the target variable.
Testing unconditional independence ($\yhat \independent \c$) between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confounder-bias in predictive modelling.
For instance, even if $\yhat \independent \c$ is false, $\yhat$ might be only marginally dependent on $\c$, due to the dependence of both on $\y$. In other words, if the target variable $\y$ is significantly associated to the confounder variable $\c$, a model that is completely blind to $\c$ (and, therefore, unbiased) might still provide outputs $\yhat$ that are significantly associated with $\c$.

@@

Instead of focusing on the unconditional independence between the confoudner and the predictions, we shall consider the \emph{conditionally independence} between $\yhat$ and $\c$ given $\y$ (written as $\yhat \independent \c | \y$) which, according to \cite{dawid1979conditional}, implies that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(\yhat|\y)\mathbb{P}(\c|\y)$. The test with this null hypothesis will be referred to as the \emph{partial confounder test}.

Alternatively, one might be also interested in testing $\yhat \independent \c | \y$. We will refer to the corresponding test as the \emph{full confounder test}.

 Conditional independence - in its general form - lies at the heart of several fundamental concepts in statistics and  plays an increasingly important role in various fields of applied statistics, particularly in biomedical applications \citep{spirtes2000causation, peters2016causal, fiedler2011mediation, candes2016panning}. Recently, \cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
 Their "no free lunch" theorem implies that, without placing some assumptions on the joint distribution of $(\y, \yhat, \c)$, conditional independence testing is effectively impossible. In other words, neither the full nor the partial confounder tests can be constructed so that - for all distributions - they provides a valid type I error control and, at the same time, a non-trivial statistical power.

@@

Even if constructing totally non-parametric confoudner tests was proven to be impossible, as shown by \cite{berrett2020conditional} and \citep{candes2016panning}, valid and powerful conditional independence tests can be constructed with inputting information about the distribution of only two (out of the three) variables.
Specifically, the conditional permutation test (CPT) of Berret and colleagues samples from a non-uniform distribution over the set of possible permutations $\pi$ of one of the variables, based on its conditional distribution of the other variable, thereby it incorporating available distributional information into permutation-based inference.

@@

As many related papers, the work of \cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is a set of predictors (features), $Y$ is the target variable and $Z$ is a potential confounder. In this setting, testing the null hypothesis $X \independent Y | Z$ aims to determine, whether the features $X$ still affect $Y$, when controlling for $Z$.
For instance, in genome-wide association studies, CPT can be used to determine whether a particular genetic variant $X$ affects a response $Y$ such as disease status or some other phenotype, even after controlling for the rest of the genome, encoded in $Z$.

@@

Within this setting, conditional independence testing and, specifically, the framework of conditional permutation testing allows investigating three different null hypotheses corresponding to the $(\y, \yhat, \c)$ triplet, as listed in Table \ref{tab:conditional-independence-cases}. Testing the null hypothesis $\y \independent \yhat | \c$ (option 1, full confoudner testing) investigates whether the model predictions are likely explainable solely with the confounder, i.e. whether the model is exclusively confounder-driven. Testing $\y \independent \c | \yhat$ (option 2) addresses the question whether the model captures all the variance in $c$ when predicting $y$. Testing the null hypothesis $\yhat \independent \c | \y$ (option 3, partial confounder testing) examines, whether the dependence of the model output on the confounder can likely be explained by the confounder's dependence on the target variable, i.e. whether there is any confounder bias in the model.

@@

Option 3, i.e. partial confounder testing is typically of interest when testing confounder bias of predictive models. Option 1, i.e. full confounder testing, might be also useful in diagnostics of predictive models, especially in the exploratory phase of model construction. Option 2 seems less appealing for model diagnostics and importantly, in this case the CPT framework does not allow constructing a test which is non-parametric on $\yhat$.

Below, CPT is adapted for partial confounder testing (option 3). The formulation of the full confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.
\subsubsection*{The partial confounder test}

@@

\begin{equation}
    c_i^{(j)} \sim Q(\cdot|y_i)
     \label{eq:cond-copy}
\end{equation}

where, $Q(.|y)$ denotes the conditional distribution of $\c$ given $\y=y$ and $j=1,\dots, m$ indexes the "copy" of $\c$ so that

$$ \c^{(j)} = (c_1^{(j)}, \dots, c_n^{(j)}) = (c_{\pi_1^{(j)}}, \dots, c_{\pi_n^{(j)}}) =  \c_{\boldsymbol{\pi}^{(j)}} $$

is a permutation of the original vector $\c = (c_1, \dots, c_n)$, with its elements reordered according to the permutation $\quad \boldsymbol{\pi} \in S_n$ where $S_n$ denote the set of all permutations on the indices $\{1,\dots,n\}$.

To ensure that Eq. \ref{eq:cond-copy} holds, the $\c_{\boldsymbol{\pi}^{(j)}}$ copies must be drawn so that:

\begin{equation}
    \label{eq-pperm}
    \mathbb{P}(\boldsymbol{ \pi }^{(j)} = \boldsymbol{ \pi } | \y,\yhat,\c) = \frac{q^n(\c_{\boldsymbol{ \pi }} | \y)}{\sum_{\boldsymbol{ \pi } ' \in S_n} q^n(\c_{\boldsymbol{ \pi } ' } | \y)}
\end{equation}

that is, according to the $q^n(\cdot|\y) := q(\cdot | y_1) \dots q(\cdot|y_n)$ product density corresponding to the conditional distribution $Q(\cdot|\y)$. Note that Eq. \ref{eq-pperm} does not necessarily assume a continuous distribution.

This mechanism creates copies $\c^{(1)}, \dots ,\c^{(m)}$ so that under the null hypothesis ($\yhat \independent \c | \y$), the triples

$$(\y,\yhat,\c), (\y, \yhat, \c^{(1)}),\dots, (\y, \yhat, \c^{(m)})$$

are all identically distributed and exchangeable, and so are the

$$T(\y,\yhat,\c), T(\y, \yhat, \c^{(1)}),\dots,T(\y, \yhat, \c^{(m)})$$

test statistics, as well.

@@

For a verification of the valid type I error control of this approach refer to Theorem 1 in \citep{berrett2020conditional}.

While the required permutations could be theoretically sampled with a simple Metropolis-Hastings algorithm that draws uniformly from $S_n$ at random, this way the acceptance ratio would be extremely low, even for moderate $n$ (except there is very low dependence of $\c$ on $\y$), resulting in slow mixing times. The partial confounder test can be, however, efficiently implemented with the parallelized pairwise Markov-Chain Monte Carlo sampler of \cite{berrett2020conditional} (Algorithm 1), that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data. The acceptance odds ratio of swapping indices $i$ and $j$ is:

\begin{equation}
\frac{ q(c_j | y_i) q(c_i | y_j)}{q(c_i | y_i) q(c_j | y_j) }
=
log \mathcal{L}(c_j | y_i) + log \mathcal{L}(c_i | y_j) - log \mathcal{L}(c_i | y_i) - log \mathcal{L}(c_j | y_j) 
\label{eq:accept-odds}
\end{equation}

In their Theorem 2, \cite{berrett2020conditional} verify that the resulting Markov Chain yields the desired stationary distribution, even if the number of steps is small.

\subsubsection*{Gaussian conditional log-likelihood}

@@

with the assumption that the residual error follows a Gaussian distribution. If we write $\boldsymbol{\mu} = \alpha + \beta \y$ and $\sigma$ is the standard deviation of the residual $\boldsymbol{e}$, then the conditional distribution of interest is given as $ (\c|\y=y_i) \sim \mathcal{N}\{\mu_i, \sigma^2\}$ and the log-likelihood, that is to be used in Eq. \ref{eq:accept-odds}, can be computed simply as the log of the corresponding probability density function:

$$ log\mathcal{L}(c_i|y_j) = - \frac{1}{2} \Big(\frac{c_i-\mu_j}{\sigma}\Big)^2 - ln(2 \pi \sigma)   $$

@@

Notably, from the three options for conditional independence-based null hypotheses enumerated in Table \ref{tab:conditional-independence-cases}, the proposed Gaussian likelihood estimation is possible for options 1 and 3, but not option 2, as the variable on which the independence is conditional must be always the predictor variable in Eq. \ref{eq:linreg}.
For options 1 and 3, the proposed simple Gaussian conditional likelihood estimation allows for deploying CPT without a large, independent sample to estimate the conditional distribution.

While, in theory, any predefined test statistic $T$ can be used, for the proposed Gaussian likelihood estimation, the python package \emph{mlconfound} implements only the two-sided variant of the tests (as on-sided testing is out-of-interest in the confoudner problem) and uses the coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic:

@@

for any $\pi \in S_n$.

@@

'Partial' and 'full' confound testing was performed with default parameters (10000 permutations, 50 Markov-chain Monte Carlo steps) as implemented in version 0.9.0 of the package \emph{'mlconfound'}. Unconditional dependence across the involved variables was investigated with conventional permutation test on the $R^2$ values, with 1000 permutations.

@@

Simulations based on normally distributed data justified that the proposed confounder tests - in line with theory - provide a valid type I error control (For detailed results, see \ref{fig:sim-normal} for the 'partial' test on numerical data and Supplementary Figures \ref{fig:sim-ccc-full}, \ref{fig:sim-ccb-partial}, \ref{fig:sim-ccb-full}, \ref{fig:sim-bbc-partial}, \ref{fig:sim-bbc-full}, \ref{fig:sim-bbb-partial} and \ref{fig:sim-bbc-full} for the partial test on categorical data and the full tests on both numeric and categorical data). 
In case of a valid test, the expected type I error rate in the null hypothesis simulation cases (first row and first column of the figures) is less then or equal to the alpha level (0.05). The upper 95\% confidence limit for the type I error rates for one simulation case (i.e. a single tile on the figures) is 0.11 (binomial confidence intervals with 100 repetitions). 
The results suggest, that the proposed tests are somewhat overly conservative if the model has a high predictive performance (type I error rates closer to 0 than expected).

Simulations with non-normal data showed that the proposed methods are remarkably robust to the violation of normality assumptions for the target variable $\y$ and the confounder $\c$ and, as expected, completely non-parametric regarding the distribution of the predictions $\yhat$). The six simulation cases of normality violation, shown on Figure \ref{fig:sim-non-normal}, involved various \emph{sinh-arcsinh} distributions, form normal (Fisher's kurtosis: 0, skewness) to severely skewed and heavy tailed (kurtosis: 42, skewness: 6).
The results show that the violation of normality - as expected - increases the type I error rate, but the increase - depending on the application - might not be severely problematic up to kurtosis and skewness values as high as 5 and 2 respectively. Furthermore, normality violation becomes less problematic in case of high predictive performance (high association between the predicted and observed values).

@@

Power estimates for the 'partial' test (with normality) are shown on Figure \ref{fig:sim-normal}. Notably, with sample sizes as large as 1000, a confounder contributing to only 0.4-0.6\% to the variance of the predictions can already be successfully detected ($w_c = 0.05 - 0.07$) with a power of 60-90\%. With a sample size of 500, a confounder contributing to only 1.4-3\% to the variance of the predictions can be detected ($w_c = 0.1 - 0.14$) with a power greater than 80\% in most of the simulation cases. A sample size of 100 yields an approximate power of 60-100\% for a confounder-variance of 6-7\% ($w_c = 0.2 - 0.21$) in most of the cases. Finally, even with a relatively low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%. If the confounder explains more than 15\% of variance, it is almost certainly detected with $n \geq 50$.

@@

Both acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p<0.001$ and $=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
The partial confounder test revealed that, the model was significantly biased both by age group and acquisition batch (both $p<0.0001$) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.084$ and $0.041$, respectively) associated to larger predicted values.

After applying confound mitigation approaches (feature regression or COMBAT) the partial confound test did not provide evidence of confounder bias anymore ($p > 0.05$ for all). Both feature regression and COMBAT decreased the predictive performance in the case of both confounders, with COMBAT providing better performances ($R^2=0.75$ and $0.76$ when applied to remove the effect of acquisition and age, respectively).

@@

Imaging center and in-scanner motion (normalized mean framewise displacement) were statistically significantly associated with ASD diagnosis ($R^2=0.019$ and $0.028$, respectively, $p<0.001$ for both, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted diagnosis with a medium effect size ($R^2=0.126$, $ROC AUC = 0.71$, $p<0.001$).
The partial confounder test revealed that the raw model was significantly biased both for age group and acquisition batch (both $p<0.0001$). Several sites (e.g. Carnegine Mellon University, University of Leuven, Social Brain Lab UMC Groningen) and lower motion being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to a lower probability for the participant to be assigned to the ASD group by the model.

Both feature regression and COMBAT seemed to significantly attenuate the effect of center, however the partial confound test still provided evidence for a significant center-bias ($p < 0.04$ and $0.01$ for feature regression and COMBAT, respectively).

When trying to mitigate the effect of in-scanner motion, feature regression obviously failed to remove the motion-bias from the model and, in fact, it introduced a paradoxical inverse dependence of the predictions on motion estimates. The partial confounder successfully detected this strong bias ($p<0.0001$). Applying COMBAT to remove the effect of motion (by using binned motion estimates), in turn, seemed to effectively mitigate motion-bias, as suggested by the partial confound test ($p=0.64$).

Both feature regression and COMBAT slightly improved the predictive performance when mitigating center-effects ($R^2=0.111$ and $0.132$ and $AUC=0.76$ and $0.75$, respectively). With both feature regression and COMBAT, the effort of mitigating motion effects happened at the cost of a drop in predictive performance ($R^2=0.02$ and $0.111$ and $AUC=0.59$ and $0.70$, respectively)

The full confounder test was highly significant ($p<0.0001$) for all models, except when regressing-out motion from the features ($p=0.09$) Indicating that this model might have been (almost) exclusively driven by motion artifacts.

% tab:unconditional-pvals
\renewcommand{\arraystretch}{1.2}
\begin{table}
\centering
\begin{tabular}{lll|ll|ll|ll|ll} 
dataset & conf. & method & $R^2_{y, c}$ & $p_{y, c}$ & $R^2_{\hat{y}, c}$ & $p_{\hat{y}, c}$ & $R^2_{\hat{y}, y}$ & $p_{\hat{y}, y}$ & partial test & full test  \\
\hline
HCP & acq.  & raw      & 0.032 & <0.001 & 0.084 & <0.001  & 0.095 & <0.001 & <0.001 & <0.001 \\
    &              & f.reg.    & & &0.003 & 0.99 & 0.066  & <0.001  & 0.99 & <0.001 \\
    &              & COMBAT    & & &0.008 & 0.75 & 0.075  & <0.001  & 0.88 & <0.001\\
\hline
    & age   & raw       & 0.011 & 0.001  & 0.041 & <0.001  & 0.095  & <0.001 & <0.001 & <0.001 \\
    &       & f.reg.    & & &0.001 & 0.3 & 0.07  & <0.001  & 0.5 & <0.001 \\
    &       & COMBAT    & & &0.004 & 0.055 & 0.076 & <0.001 & 0.12 & <0.001 \\
\hline
ABIDE   & center   & raw       & 0.019  & <0.001 &  0.169 & <0.001& 0.126     & <0.001 & <0.001 & <0.001 \\
        &          & f.reg.    &  & &  0.042 & 0.007 & 0.173     & <0.001 & 0.04 & <0.001 \\
        &          & COMBAT    &  & &  0.05 & 0.001 & 0.17     & <0.001 & 0.01 & <0.001 \\
\hline
        & motion   & raw       & 0.028 & <0.001 & 0.111    &  <0.001 & 0.126    & <0.001 & <0.001& <0.001 \\
        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & <0.001 & 0.09 \\
        &          & COMBAT    & & &  0.002 & 0.202 & 0.111     & <0.001 & 0.64 & <0.001 \\

@@

Confounder-bias can significantly harm the generalizability and biomedical validity of predictive models. Testing confounder-bias of predictive models can be achieved by considering the "confounder-problem" as assessing \emph{conditional independence} between the predicted values and either the target or the confounder variable. These two scenarios result in two different test, testing whether the model is partially or fully driven by the given confounder.
Importantly, in predictive modelling, the model predictions are often follow a non-normal distribution, thus any test for confounder bias must be, at least partly, non-parametric.
However, as shown by \citet{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
Here I propose the use of the conditional permutation test (CPT) of \cite{berrett2020conditional} to test confounder bias in predictive models. Adapting this approach for the confounder testing problem makes it possible to place assumptions about the joint distribution of only two out of the three variables involved; that is, it allows constructing both the partial and full confounder tests so that no assumption is made about the distribution of the model output. 
The tests are made available in the python package \emph{mlconfound} which builds upon a parallelized, computationally effective implementation of the Markov-chain Monte-Carlo based parallel pairwise sampler of \cite{berrett2020conditional}. It utilizes, coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confounder is normal. The partial and full confoudner tests available in \emph{mlconfound} can be uniformly applied for any predictive model, given that realistic (e.g. cross-validated) model predictions can be obtained.

@@

Given the power and the robustness of the proposed tests, they might become a useful tool in diagnosing predictive models in a wide spectrum of applications.
